---
title: "Sensor Signal Analysis for Predicting Semiconductor Manufacturing Defects"
subtitle: Paul J Anderson
output:
  pdf_document: default
html_notebook: default
html_document:
  df_print: paged
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
# Core tidyverse for data wrangling & plotting
library(tidyverse)
library(corrplot)
library(zoo)
library(here)

# Modeling
library(caret)
library(pROC)
library(glmnet)
```

# Abstract
Semiconductor manufacturing yield depends on early failure detection.
This study analyzes the SECOM dataset--1,567 observations from a
semiconductor fabrication line, each characterized by 590 anonymized
sensor measurements and a binary pass/fail label. The data present
three compounding challenges: severe class imbalance (~93% pass),
substantial sensor missingness, and a high-dimensional feature space
relative to the number of observed failures (n = 99). We address two
research questions: which sensors are most strongly associated with
product failure, and can failure be predicted reliably from sensor
readings alone? Following a principled preprocessing pipeline and
chronological holdout evaluation, we fit an elastic net-penalized
logistic regression with alpha and lambda selected by cross-validated
AUC. A single sensor--Sensor 59--is selected by the final model
and independently confirmed by Wilcoxon rank-sum test (p = 2.30 x $10^{-8}$),
achieving a test AUC of 0.72. The convergent identification of one
dominant process signal across two independent methods is the central
finding, and suggests Sensor 59 as a priority candidate for process
engineer investigation.

# Introduction
In semiconductor manufacturing, product yield is a critical performance
metric. A single fabrication line produces thousands of units across
hundreds of tightly controlled process steps--including
photolithography, chemical mechanical planarization (CMP), physical
and chemical vapor deposition (PVD/CVD), etch, and metrology--each
monitored by an array of sensors. These sensors capture parameters
such as electrical characteristics (leakage current, sheet resistance,
threshold voltage), physical layer properties (film thickness,
planarity, step height), process conditions (chamber pressure, gas
flow rates, radio-frequency power), and optical measurements (critical
dimension, overlay error). When a unit fails quality inspection,
identifying which upstream process conditions contributed to that
failure is both technically difficult and economically important--
catching failures early reduces waste, scrap costs, and
time-to-diagnosis for process engineers.

The SECOM dataset (UCI Machine Learning Repository) provides a
concrete case for this problem. Collected from a real fabrication
line over eleven months in 2008, it contains 1,567 observations and
590 anonymized sensor measurements per unit, along with a binary
pass/fail outcome. Although sensor identities are anonymized, the
measurement types likely span electrical test results, deposition
uniformity metrics, planarization endpoint signals, and etch rate
monitors--making the dataset broadly representative of real fab-line
data. Its challenges are equally representative: a failure rate of
approximately 7% creates severe class imbalance; substantial missing
data across many sensors is consistent with intermittent tool downtime
or recipe-specific measurement steps; and the feature space is
high-dimensional relative to the number of observed failures,
with an events-per-variable ratio well below the recommended minimum
of 10.

This analysis pursues two goals. The first is explanatory: to
identify which sensors show meaningful differences in readings between
passed and failed units, providing insight into which process
conditions are most associated with failure. The second is predictive:
to develop a classification model capable of forecasting product
failure from sensor readings, evaluated honestly against the
constraints imposed by class imbalance and limited failure events.

# Data Processing

We apply a preprocessing pipeline to improve data quality and reduce dimensionality. First, sensors with more than 150 missing values (approximately 10% of observations) are removed, as many measurements are only collected for specific process recipes. Next, rows containing any remaining missing values are dropped to produce a fully observed dataset. The binary pass/fail outcome is then recoded from {-1, +1} to {0, 1} for compatibility with logistic regression. Sensors with zero variance are removed because they provide no predictive information. Finally, variance-based trimming is applied to address extreme variability across sensors: features outside the 35th–80th percentile of variance are excluded, since very low-variance sensors contribute little information and extremely high variances are likely driven by outliers rather than meaningful process signals. After preprocessing, the final analysis dataset contains 1,393 observations and 190 sensor features, with a failure rate of 7.1% (99 failures and 1,294 passes).

```{r}
# chunk 1--load
secom <- read_csv(here("defect_report/uci-secom.csv"))

# chunk 2--clean
secom_clean <- secom |>
  select(where(~ sum(is.na(.)) <= 150)) |>
  drop_na() |>
  mutate(`Pass/Fail` = ifelse(`Pass/Fail` == -1, 0, 1)) |>
  rename(Label = `Pass/Fail`) |>
  select(where(~ var(.x, na.rm = TRUE) != 0)) |>
  arrange(Time) |>
  mutate(obs_number = row_number())

# chunk 3--variance filter (THIS IS MISSING)
variances       <- sapply(secom_clean, function(x) var(x, na.rm = TRUE))
lower_threshold <- quantile(variances, 0.35)
upper_threshold <- quantile(variances, 0.80)
keep_cols       <- variances > lower_threshold & variances < upper_threshold
data_clean      <- secom_clean[, keep_cols]

cat("data_clean:", nrow(data_clean), "obs,", 
    ncol(data_clean), "features\n")
cat("Failures:", sum(data_clean$Label == 1), "\n")
```

```{r}
ggplot(secom_clean, aes(x = factor(Label,
                       labels = c("Pass (0)", "Fail (1)")))) +
  geom_bar(fill = c("grey", "grey"), width = 0.5) +
  labs(title = "Distribution of Product Outcomes",
       x = "Outcome", y = "Count") +
  theme_classic()
```


The dataset is severely imbalanced: 1,294 units passed and only 99
failed. This ratio has direct consequences for modeling. A naive
classifier that predicts "pass" for every unit achieves 92.9% accuracy
while detecting zero failures. Standard model accuracy is therefore
uninformative here; we rely instead on AUC, sensitivity, and balanced
accuracy instead.

Observations span January 8 to December 10, 2008. Manufacturing
processes evolve over time, tool drift, recipe changes, and maintenance
events can shift the failure rate non-stationarily. We examine failure
rate over time using a rolling 50-observation window.

```{r}
secom_clean |>
  arrange(obs_number) |>
  mutate(failure           = as.numeric(Label == 1),
         rolling_fail_rate = zoo::rollmean(failure, k = 50,
                                           fill = NA, align = "right")) |>
  ggplot(aes(x = obs_number, y = rolling_fail_rate)) +
  geom_line(color = "firebrick") +
  labs(title = "Rolling Failure Rate Over Observations (k = 50)",
       x = "Observation Number",
       y = "Rolling Failure Rate") +
  theme_minimal()
```

The spike near observation 450 (timestamp: 2008-08-09 00:32 UTC) is
the most prominent feature of the temporal profile. Its sharpness
suggests a discrete process event rather than gradual drift. This
non-stationarity means observations are not fully exchangeable over
time--a violation of the independence assumption in logistic
regression. Standard k-fold cross-validation does not account for
this. Chronological holdout validation would be more appropriate and
is noted as a priority for future work.

```{r}
cor_mat <- cor(data_clean[, sapply(data_clean, is.numeric)])
corrplot(cor_mat, method  = "color",
         order  = "hclust",  # groups correlated sensors together
         tl.pos = "n",
         title  = "Sensor Correlation--All Variables (clustered)",
         mar    = c(0, 0, 2, 0))
```

The correlation matrix reveals pronounced block structure among the
among all 190 sensors, with several clusters of highly correlated
variables. This is consistent with sensors grouped by process stage —
for example, multiple planarity or film thickness measurements taken
at the same tool step would be expected to co-vary. For modeling,
this correlation structure has an important implication: standard
logistic regression becomes unstable when predictors are highly
collinear. LASSO penalization addresses this directly by selecting
one representative from correlated clusters and shrinking the rest
toward zero, which is a key justification for the modeling approach
taken in Section 3.

# Methods & Models
We use a chronological holdout split rather than random partitioning.
Random splits allow future observations to leak into training, which
artificially inflates performance estimates when the failure process
is non-stationary--as is clearly the case here given the pronounced
spike near observation 450 (2008-08-09 00:32 UTC).

We initially explored splitting at the spike, training on the
pre-anomaly period (n=450, 42 failures) and testing on the
post-anomaly period (n=943, 57 failures). This framing asks a
meaningful question--can pre-anomaly sensor patterns generalize
to a post-anomaly process--but yielded insufficient training
failures for stable model estimation.

We therefore apply an 80/20 chronological split, assigning the
first 1,114 observations to training (86 failures, 7.7%) and the
remaining 279 to the test set (13 failures, 4.7%). Temporal order
is preserved throughout, preventing future data leakage while
retaining enough failures for meaningful model estimation.

```{r chronological-split, echo=FALSE}
data_clean <- data_clean |> mutate(obs_number = row_number())

n         <- nrow(data_clean)
train_data <- data_clean |> filter(obs_number <= floor(0.8 * n))
test_data  <- data_clean |> filter(obs_number > floor(0.8 * n))

cat("Training:", nrow(train_data), "obs,",
    sum(train_data$Label == 1), "failures\n")
cat("Test:    ", nrow(test_data), "obs,",
    sum(test_data$Label == 1), "failures\n")
```

We fit an elastic net-penalized logistic regression using `glmnet`,
combining L1 and L2 penalties through a mixing parameter $\alpha$:

$$\text{Penalty} = \alpha \|\beta\|_1 + \frac{(1-\alpha)}{2}\|\beta\|_2^2$$

Pure LASSO ($\alpha$ = 1) tends to arbitrarily discard sensors from
correlated clusters, losing signal from related process stages.
Pure ridge ($\alpha$ = 0) retains all features, preventing the sparse
sensor identification needed to answer the explanatory question.
Elastic net balances both. With 86 training failures across 190
features, EPV is well below 1, making penalization essential to
control overfitting.

$\alpha$ is selected via grid search over {0.25, 0.50, 0.75, 1.00},
with $\lambda$ tuned by cross-validated AUC for each $\alpha$ value. The optimal
threshold is selected using Youden's J statistic on the training
ROC curve and applied to the test set to avoid data leakage.

Feature importance is assessed through elastic net coefficient
magnitudes and Wilcoxon rank-sum tests on all 190 sensors, with
a Bonferroni correction applied to control family-wise error rate.

```{r modeling, include=FALSE}
x_train <- as.matrix(train_data[, setdiff(names(train_data), "Label")])
y_train <- train_data$Label
x_test  <- as.matrix(test_data[, setdiff(names(test_data), "Label")])
y_test  <- test_data$Label

# alpha grid search
set.seed(1187)
alpha_grid  <- c(0.25, 0.50, 0.75, 1.00)
auc_results <- sapply(alpha_grid, function(a) {
  fit <- cv.glmnet(x_train, y_train, alpha = a,
                   family = "binomial", type.measure = "auc")
  max(fit$cvm)
})

best_alpha <- alpha_grid[which.max(auc_results)]
best_fit   <- cv.glmnet(x_train, y_train, alpha = best_alpha,
                        family = "binomial", type.measure = "auc")

# threshold from training ROC only
train_prob  <- predict(best_fit, newx = x_train,
                       s = "lambda.1se", type = "response")
roc_train   <- roc(y_train, as.vector(train_prob))
best_thresh <- as.numeric(coords(roc_train, "best",
                                 best.method = "youden",
                                 transpose   = FALSE)["threshold"])

# test set evaluation
test_prob  <- predict(best_fit, newx = x_test,
                      s = "lambda.1se", type = "response")
roc_test   <- roc(y_test, as.vector(test_prob))
test_class <- ifelse(as.vector(test_prob) >= best_thresh, 1, 0)
conf       <- confusionMatrix(factor(test_class), factor(y_test),
                               positive = "1")

cat("Best alpha:", best_alpha, "\n")
cat("Sensors selected:", sum(coef(best_fit, s = "lambda.1se")[-1] != 0), "\n")
cat("Optimal threshold:", round(best_thresh, 4), "\n")
cat("Test AUC:", round(auc(roc_test), 4), "\n")
#print(conf)
```
# Results

Grid search over $\alpha$ $\epsilon$ {0.25, 0.50, 0.75, 1.00} selected $\alpha$ = 0.25,
reflecting a predominantly ridge blend with modest L1 sparsity
consistent with the block correlation structure observed among
sensors. At lambda.1se, the model retains a single sensor, Sensor 59,
with a positive coefficient of 0.0037. Elevated Sensor 59 readings
are associated with increased failure probability. The intercept of
-2.50 reflects the low baseline failure rate in the training period.

```{r}
knitr::kable(
  data.frame(
    Term        = c("Intercept", "Sensor 59"),
    Coefficient = c(-2.4958, 0.0037)
  ),
  caption = "Elastic net coefficients at lambda.1se (alpha = 0.25).",
  digits  = 4
)
```

The model achieves a test AUC of 0.72 on the chronological holdout
set. At the Youden-optimal threshold of 0.077, the model correctly
identifies 4 of 13 failures (sensitivity = 0.31) while flagging only
14 of 266 passing units as false alarms (specificity = 0.95).
Balanced accuracy of 0.63 and kappa of 0.22 both reflect modest but
meaningful discrimination above chance. Notably, the less regularized
lambda.min model--retaining 68 sensors--achieves a lower test AUC
of 0.66, confirming that additional sensors introduce noise rather
than signal under these low EPV conditions.

```{r}
knitr::kable(
  data.frame(
    Metric = c("Test AUC", "Sensitivity", "Specificity",
               "Balanced Accuracy", "Precision", "Kappa"),
    Value  = c(0.7169, 0.308, 0.947, 0.628, 0.222, 0.216)
  ),
  caption = "Model performance on chronological holdout test set (n=279, 13 failures)."
)
```

Wilcoxon rank-sum tests across all 190 sensors identify six sensors
with statistically significant distributional differences between
passed and failed units after Bonferroni correction ($\alpha$ = 0.05/190
~ 0.00026): Sensors 59, 33, 129, 28, 477, and 205.

Sensor 59 is the dominant signal by a wide margin--its corrected
p-value (2.30 × $10^{-8}$) is three orders of magnitude below the next
most significant sensor (Sensor 33, corrected p = 0.010). The five
remaining sensors are near the correction threshold and should be
interpreted cautiously given the limited number of failures. The
independent convergence of elastic net and Wilcoxon screening on
Sensor 59--through entirely different analytical frameworks--is
the central finding of this analysis.

```{r wilcoxon-all, echo=FALSE}
sensor_cols <- setdiff(names(data_clean), "Label")

wilcox_results <- data.frame(
  Sensor  = sensor_cols,
  p_value = sapply(sensor_cols, function(s) {
    wilcox.test(data_clean[[s]] ~ data_clean$Label)$p.value
  })
) |>
  arrange(p_value) |>
  mutate(
    p_adjusted  = p.adjust(p_value, method = "bonferroni"),
    Significant = p_adjusted < 0.05
  )

# summary
cat("Sensors significant after Bonferroni correction:",
    sum(wilcox_results$Significant), "of", nrow(wilcox_results), "\n")

# top 15 table
knitr::kable(
  head(wilcox_results, 15) |>
    mutate(p_value    = formatC(p_value,    format = "e", digits = 2),
           p_adjusted = formatC(p_adjusted, format = "e", digits = 2)) |>
    rename(Sensor = Sensor,
           `p-value` = p_value,
           `Bonferroni p` = p_adjusted,
           `Significant` = Significant),
  caption = "Top 15 sensors by Wilcoxon rank-sum test (all 190 sensors tested, Bonferroni corrected)."
)
```

```{r sig-sensors-plot, echo=FALSE, fig.cap="Distributions of the six Bonferroni-significant sensors by product outcome. Sensor 59 shows the most pronounced separation."}
sig_sensors <- wilcox_results |>
  filter(Significant) |>
  pull(Sensor)

data_clean |>
  mutate(Label = factor(Label, labels = c("Pass", "Fail"))) |>
  select(Label, all_of(sig_sensors)) |>
  pivot_longer(-Label, names_to = "Sensor", values_to = "Reading") |>
  ggplot(aes(x = Label, y = Reading, fill = Label)) +
  geom_boxplot(outlier.alpha = 0.3) +
  scale_fill_manual(values = c("Pass" = "steelblue", "Fail" = "tomato")) +
  facet_wrap(~ Sensor, scales = "free_y") +
  labs(title = "Significant Sensor Readings by Product Outcome",
       x = "Outcome", y = "Sensor Reading") +
  theme_classic() +
  theme(legend.position = "none")
```

# Discussion

Sensor 59 is the dominant signal associated with product failure, identified by both elastic net selection and Wilcoxon testing (corrected p = 2.30 × $10^{-8}$). It is the most significant feature by several orders of magnitude and has a positive coefficient, indicating higher values are linked to increased failure risk. Five additional sensors (33, 129, 28, 477, 205) show weaker signals after correction. Without metadata, the physical meaning of Sensor 59 is unknown, but its consistency across methods makes it the primary candidate for process investigation.

Predictively, a single-sensor elastic net model achieves AUC = 0.72 on a chronological test set. A larger 68-sensor model performs worse (AUC = 0.66), likely due to low event density and added noise. Sensitivity is 0.31 at the Youden threshold, meaning most failures are missed, which is problematic in a high-cost escape setting. Improving recall would substantially increase false positives given the 4.7% failure rate.

The main limitations of the analysis are driven by data scarcity, temporal instability, and limited interpretability. With only 13 failures in the test set, estimates of sensitivity and AUC are inherently unstable and subject to high variance. In addition, the process appears to exhibit temporal drift, as evidenced by differing failure rates between the training period (7.7%) and the test period (4.7%), and a large spike in failures at the 450 mark, which may effect generalization. The lack of semantic information about the sensors further constrains interpretation, since Sensor 59 cannot be linked to a specific physical process without context. Finally, the use of variance-based filtering may have removed potentially informative sensors, introducing additional uncertainty into feature selection.

Given that a single sensor dominates both the predictive and
explanatory results, the most impactful next step is identifying
the physical process parameter Sensor 59 measures and investigating
whether its elevated readings correspond to a known failure mode —
tool drift, recipe deviation, or incoming material variation.

Statistically, an anomaly detection approach trained solely on
Sensor 59 readings from passing units--such as a control chart
or one-class classifier--may be more appropriate and interpretable
than a full classification model, and would generalize more naturally
to the low-failure-rate manufacturing environment.

Acquiring additional failure data or extending the observation window
would improve EPV and enable the spike-based chronological split,
which remains a more scientifically motivated evaluation than the
80/20 cut. With sufficient failures, time-series cross-validation
may provide more reliable performance estimates than a single
holdout period.
