---
title: "Evaluating the Effects of Habitat Designation and Climate on Spring-run Chinook Salmon Return Rates in Upper Columbia River Basin"
author: "Paul J Anderson"
date: "`r Sys.Date()`"
output: # bookdown::pdf_document2 <- try this for figure numbering
  pdf_document:
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_knit$set(root.dir = "C:\\Users\\paulande\\OneDrive - Oregon State University\\Classes\\2026\\Spring\\ST595_exec_sum\\salmon_report")
# sets root directory for the knitting process

# packages
library(ggplot2)
library(bookdown)
library(GGally)
library(tidyr)
library(knitr)
library(broom)
library(backports)
library(faraway)
library(MASS)
library(reshape)
library(ggExtra)
library(broom)
library(car)
library(dplyr)
library(reticulate)
library(tinytex)
library(kableExtra)
library(here) # using here to replace the absolute path with relative
```

```{r}
# load in data
# unscrubbed data
fish_data <- read.csv(here("salmon_report", "fish_data.csv"))
# climatic_data <- read.csv("climatic_variables_datapull.csv")
# storm_data <- read.csv("combined_storm_events.csv")

# post dataprep data
dataprep_data <- read.csv(here("salmon_report", "final_data_post_dataprep.csv"))
```

```{r}
# scrubbed data
final_data <- read.csv(here("salmon_report", "imputed_dataset.csv"))
```

## Introduction

Decreases in fish stocks have cultural, social, economic, and environmental impacts throughout the western United States. Tracking changes in fish stocks provides vital data for the conservation and protection of this declining resource. As part of fish stock management, various state and federal agencies monitor oceanic and river conditions to determine catch limits. The National Oceanic and Atmospheric Administration (NOAA) cites eight ocean indicators critical for monitoring the health of salmon fish stocks along the western United States.

Previous studies used one of these ocean indicators, sea surface temperatures (SST), and examined its influence on the smolt-to-adult return ratio (SAR) of an important group of salmonids: the Upper Columbia spring-run Chinook salmon. While significant relationships were present, those analyses had low statistical power. This study expands the timeframes by splitting the data by the year in which the ESU received critical habitat designation (September 2005), creating before and after designation time periods. These two time periods are investigated by collecting mean SAR and juvenile returns for the ESU alongside the explanatory variables of mean SST and max SST for the California Current between 1994 and 2024. Explanatory variables of copepod species richness and eutrophication due to flooding events along the coastal range of the California Current were investigated but reserved for a future study given the additional time required for data processing.

Further study is needed to address several key questions in this project. Specifically, this report aims to determine:

- Whether there is a significant difference in the mean Smolt to Adult Return (SAR) and Chinook Juvenile catches between the pre- and post-listing periods of critical habitat for the Spring-run, Upper Columbia Basin Chinook Salmon.

- Whether there is a significant difference in environmental variables — including mean Sea Surface Temperature (meanSST) and maximum Sea Surface Temperature (maxSST) between the pre and post-listing periods.

- Which explanatory environmental variables are most important in describing the null hypothesis ($H_0$) of no difference in the response variable (mean SAR, Upper Columbia Basin Chinook Salmon) between the pre- and post-listing periods of critical habitat.

## Methods

### Data

Fish stock, juvenile return, oceanographic, and climatic data were compiled from multiple federal and academic sources. Fish stock survival data were obtained from the Columbia Basin Research Center's Data Access in Real Time (DART) system using the SAR survival API with settings for Bonneville (All) to Bonneville Adult returns, Chinook Upper Columbia River Spring ESU, exclusion of zero-year adult detections, aggregated by release basin, and pooled-by-year outputs. Juvenile return counts were retrieved from the same DART resource using parallel filters applied to juvenile outmigrant trap data for the Upper Columbia Basin Chinook Spring ESU spanning 1994–2024. Sea surface temperature (SST) data, including annual mean and maximum values for the California Current system, were sourced from NOAA's Northwest Fisheries Science Center, spatially filtered to the California Current extent and temporally aggregated by year. Additional stochastic climatic variables were obtained from NOAA's National Center for Environmental Information; data from 1994–2024 were downloaded, decompressed, and merged using Python's pandas and glob libraries.

Fish data were scrubbed emulating the process completed by Team Nexus (1) including the full temporal range of interest and retaining the field juvenile returns. Climatic data were scrubbed using two recipe dataflows in Dataprep to process out whitespace, convert fields to integer datatype, remove unnecessary fields, add a field for year, add a field for before and after critical habitat designation Environmental Significant Unit (ESU), calculate average for SST, calculate maximum for SST, and sort by year. The second recipe in our flow filtered rows by year and retained mode of ESU, average from mean_SST, and maximum from max_SST. Fish data was then uploaded to a Google Cloud Storage bucket. A third recipe was used in our dataflow to create a join between the scrubbed climatic variables and fish data. In the interest of linear modeling, we then attempted to use the fill feature to use linear imputation for the missing values of our response variable (mean_SAR, juvenile_returns). Statistically, imputing response variables is generally discouraged; however, given the extent of missing data and our interest in exploring imputation methods as part of this analysis, we elected to proceed rather than substantially reducing the available sample size. We determined that a K‑Nearest Neighbors (KNN) approach using mean values of the nearest neighbors was appropriate for this dataset and successfully used it to impute the mean_SAR and juvenile_return fields. Imputation was performed in Python using the scikit‑learn library, with values derived from the nearest neighbors in the Dataprep-processed dataset, following established documentation (5). Data processing and hypothesis testing were conducted in Python and R, with visualization and model fitting performed locally.

### Exploratory Data Analysis

Initial evaluation of the cleaned dataset confirmed 31 annual observations spanning 1994–2024 with no missing values and correct data types across all six fields. The ESU designation variable reflects an uneven group split, with approximately 12 pre-designation and 19 post-designation years. Visual inspection of faceted boxplots (Figure 1) revealed unequal variance in juvenile return, max SST, and mean SAR, with post-ESU groups showing notably wider interquartile ranges and longer whiskers. Mean_SST showed comparatively similar spread between groups. Visual assessment of Q-Q plots (Figure 2) and Shapiro-Wilk tests (Table 1) failed to reject normality for all variables except mean SAR in the pre-ESU group (W = 0.845, p = 0.032). The compressed variance and departure from normality observed in mean SAR and juvenile return during the pre-ESU period are likely attributable to KNN imputation of missing values, which may understate true variability in that group. Based on the visual variance assessment, unequal group sizes, and normality results, Welch's t-test was selected over Student's t-test for all comparisons. Results for mean SAR should be interpreted with caution given the normality violation in the pre-ESU group. Autocorrelation function (ACF) plots (Figure 3) revealed no significant autocorrelation in mean SAR, juvenile return, or Mean_SST. Max SST exhibited a marginal lag-2 spike that remained within confidence bounds, noted as a minor limitation but insufficient to violate the independence assumption.


```{r, include=FALSE}
# include false this output
str(final_data)
summary(final_data)
colSums(is.na(final_data))
```

```{r, include=FALSE}
final_data
```

```{r EDA-1, echo = FALSE, fig.cap = "Figure 1. Visual variance check."}
# reshape to long format for faceting
fish_long <- final_data |>
  mutate(esu = factor(esu, labels = c("Pre-ESU", "Post-ESU"))) |>
  pivot_longer(
    cols = c(mean_SAR, juvenile_return, max_SST, Mean_SST),
    names_to = "variable",
    values_to = "value"
  )

# plot
ggplot(fish_long, aes(x = esu, y = value, fill = esu)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_fill_manual(values = c("#2171b5", "#cb181d")) +
  theme_minimal(base_size = 13)
```
```{r EDA-2, echo = FALSE, fig.cap = "Figure 2. Q-Q plots, visual check of normality."}
# QQ plots
ggplot(fish_long, aes(sample = value, color = esu)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ variable, scales = "free") +
  scale_color_manual(values = c("#2171b5", "#cb181d")) +
  theme_minimal(base_size = 13)
```
```{r EDA-3, echo = FALSE, tbl.cap = "Table 1. Shapiro-wilk test formal test of normality by group."}
fish_long |>
  group_by(variable, esu) |>
  summarize(
    W       = shapiro.test(value)$statistic,
    p_value = shapiro.test(value)$p.value,
    .groups = "drop"
  ) |>
  arrange(variable, esu) |>
  kable(
    col.names = c("Variable", "ESU Group", "W", "p-value"),
    caption   = "Table X. Shapiro-Wilk Tests of Normality by Variable and ESU Group"
  )
```

```{r EDA-4, echo = FALSE, fig.cap = "Figure 3. ACF plots, visual check of independence."}
par(mfrow = c(2, 2))
acf(final_data$mean_SAR,        main = "ACF - mean SAR")
acf(final_data$juvenile_return, main = "ACF - Juvenile Return")
acf(final_data$max_SST,         main = "ACF - Max SST")
acf(final_data$Mean_SST,        main = "ACF - Mean SST")
```

### Models

#### Question 1. 
Is there a significant difference in the mean Smolt to Adult Return (SAR) and Chinook Juvenile catches between the pre and post listing periods of critical habitat for the Spring-run, Upper Columbia Basin Chinook Salmon? We pose these questions with the following testable hypotheses.


$$
H_0: meanSAR_{pre} = meanSAR_{post}
$$
$$
H_A: meanSAR_{pre} \neq meanSAR_{post}
$$

$$
H_0: Juveniles_{pre} = Juveniles_{post}
$$

$$
H_A: Juveniles_{pre} \neq Juveniles_{post}
$$

Welch's two-sample t-tests were conducted to evaluate whether mean SAR and juvenile return differed significantly between pre and post-ESU designation periods. No statistically significant difference was found in mean SAR between pre-ESU and post-ESU periods (p-value = 0.072), though the result approached significance at $\alpha$ = 0.05. Juvenile return similarly showed no significant difference between pre-ESU and post-ESU periods (p-value = 0.288). The null hypothesis was retained for both variables. It should be noted that the marginal SAR result (p = 0.072) may warrant further investigation with a larger sample, and that both results should be interpreted cautiously given the KNN imputation of missing values in the pre-ESU period and the low statistical power associated with the available sample size of n = 31. (Table 2)

```{r mod-1a, include=FALSE}
# Q1A: Smolt to Adult Return (SAR)
sar_test <- t.test(mean_SAR ~ esu, data = final_data, var.equal = FALSE)
sar_test

# Q1B: Juvenile Return
juv_test <- t.test(juvenile_return ~ esu, data = final_data, var.equal = FALSE)
juv_test
```

```{r mod-1b, echo = FALSE, tbl.cap = "Table 2. Welch Two-Sample t-test Results for Question 1."}
# create a tidy table
bind_rows(
  tidy(sar_test) |> mutate(metric = "SAR"),
  tidy(juv_test) |> mutate(metric = "Juvenile Return")
) |>
transmute(
    Variable        = metric,
    `Mean Pre-ESU`  = estimate1,
    `Mean Post-ESU` = estimate2,
    `t-statistic`   = statistic,
    df              = parameter,
    `p-value`       = p.value,
    `CI Low`        = conf.low,
    `CI High`       = conf.high
) |>
  kable(digits = 3, caption = "Table 2. Welch Two-Sample t-test Results",
    booktabs = TRUE) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )

```


#### Question 2.
Is there a significant difference in environmental variables such as mean Sea Surface Temperature (mean_SST) and maximum Sea Surface Temperature (max_SST) between the pre and post listing periods of critical habitat for the Spring-run, Upper Columbia Basin Chinook Salmon? We post this question poses the following testable hypotheses.

$$
H_0: meanSST_{pre} = meanSST_{post}
$$

$$
H_A: meanSST_{pre} \neq meanSST_{post}
$$

$$
H_0: maxSST_{pre} = maxSST_{post}
$$

$$
H_A: maxSST_{pre} \neq maxSST_{post}
$$

Welch's two-sample t-tests were conducted to evaluate whether mean SST and max SST differed significantly between pre- and post-ESU designation periods. No significant difference was found in mean SST between pre-ESU and post-ESU periods (p-value = 0.554). Max SST however showed a statistically significant difference between pre-ESU ($\bar{X}$ = 22.366*C) and post-ESU ($\bar{X}$ = 23.424°C) periods (t = -2.641, df = 27.939, p = 0.013), with post-designation years showing notably warmer maximum sea surface temperatures. The null hypothesis was retained for mean SST and rejected for max SST. The significant max SST result is notable as it suggests measurable warming in peak ocean temperatures coinciding with the post-designation period, though the observational nature of this analysis precludes causal inference. (Table 3)

```{r mod-2a, include=FALSE}
# Q2A: Mean SST
mSST_test <- t.test(Mean_SST ~ esu, data = final_data, var.equal = FALSE)
mSST_test

# Q2B: Max SST
maxSST_test <- t.test(max_SST ~ esu, data = final_data, var.equal = FALSE)
maxSST_test
```


```{r mod-2b, echo = FALSE, tbl.cap = "Table 3. Welch Two-Sample t-test Results for Question 2."}
# create a tidy table
bind_rows(
  tidy(t.test(Mean_SST ~ esu, data = final_data, var.equal = FALSE)) |> mutate(variable = "Mean_SST"),
  tidy(t.test(max_SST ~ esu, data = final_data, var.equal = FALSE)) |> mutate(variable = "max_SST")
) |>
  select(variable, estimate1, estimate2, statistic, parameter, p.value, conf.low, conf.high) |>
  mutate(across(where(is.numeric), ~ round(., 3))) |>
  rename(
    Variable        = variable,
    `Mean Pre-ESU`  = estimate1,
    `Mean Post-ESU` = estimate2,
    `t-statistic`   = statistic,
    df              = parameter,
    `p-value`       = p.value,
    `CI Low`        = conf.low,
    `CI High`       = conf.high
  ) |>
  kable(
    caption = "Table 3. Welch's T-test Results for Question 2",
    booktabs = TRUE) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )
```

#### Question 3. 
Does sea surface temperature explain variance in mean Smolt to Adult Return (SAR) beyond that accounted for by ESU designation alone?

Model 1: mean SAR explained by ESU designation

$$
\text{meanSAR} = \beta_0 + \beta_1 \,\text{ESU}
$$
$$
H_0: \beta_1 = 0
$$

$$
H_A: \beta_1 \neq 0
$$

Model 2: mean SAR explained by max SST and ESU designation

$$
\text{meanSAR} = \beta_0 + \beta_1 \,\text{MaxSST} + \beta_2 \,\text{ESU}
$$
$$
H_0: \beta_1 = 0
$$
$$
H_A: \beta_1 \neq 0
$$
Model 3: mean SAR explained by Mean SST and ESU designation
$$
\text{meanSAR} = \beta_0 + \beta_1 \,\text{MeanSST} + \beta_2 \,\text{ESU}
$$
$$
H_0: \beta_1 = 0
$$
$$
H_A: \beta_1 \neq 0
$$
After fitting the models and prior to model interpretation, regression assumptions were evaluated for all three models. Scatterplots (Figure 4) confirmed a reasonable negative linear relationship between mean SAR and max SST in M2 and a weaker negative trend with Mean SST in M3, satisfying the linearity assumption. Residuals vs. fitted plots (Figure 5) revealed randomly scattered residuals around zero for M1 and M2, supporting homoscedasticity. M3 exhibited a mild U-shaped residual trend suggesting minor heteroscedasticity, noted as a limitation. Q-Q plots (Figure 6) and Shapiro-Wilk tests (Table 6) failed to reject normality of residuals for all three models — M1 (W = 0.972, p = 0.578), M2 (W = 0.974, p = 0.648), and M3 (W = 0.985, p = 0.936). VIF values for M2 and M3 (Table 7) were well below the conventional threshold of 5, indicating no multicollinearity concern. Observation 30 appeared as a consistent outlier across all three models and is noted as a potential influence point. Overall, assumptions were satisfactorily met for M1 and M2, with M3 warranting cautious interpretation given the mild heteroscedasticity in its residuals.

```{r mod-3a, echo = FALSE}
# Model 1: Baseline
m1 <- lm(mean_SAR ~ esu, data = final_data)

# Model 2: max SST + ESU
m2 <- lm(mean_SAR ~ max_SST + esu, data = final_data)

# Model 3: Mean SST + ESU
m3 <- lm(mean_SAR ~ Mean_SST + esu, data = final_data)
```



```{r mod-3b, echo = FALSE, fig.cap = "Figure 4. Scatterplot of mean SAR vs max SST for Linearity."}
par(mfrow = c(1, 3))
plot(final_data$esu, final_data$mean_SAR,
     main = "M1: mean SAR vs ESU",
     xlab = "ESU", ylab = "mean SAR")

plot(final_data$max_SST, final_data$mean_SAR,
     main = "M2: mean SAR vs max SST",
     xlab = "max SST (°C)", ylab = "mean SAR")

plot(final_data$Mean_SST, final_data$mean_SAR,
     main = "M3: mean SAR vs Mean SST",
     xlab = "Mean SST", ylab = "mean SAR")
```

```{r mod-3c, echo = FALSE, fig.cap = "Figure 5. Residuals vs. fitted plot for Homoscedasticity."}
par(mfrow = c(1, 3))
plot(m1, which = 1, main = "M1 Residuals vs Fitted")
plot(m2, which = 1, main = "M2 Residuals vs Fitted")
plot(m3, which = 1, main = "M3 Residuals vs Fitted")
```
```{r mod-3d, echo = FALSE, fig.cap = "Figure 6. Shapiro-Wilk for Normality of residuals."}
par(mfrow = c(1, 3))
plot(m1, which = 2, main = "M1 Q-Q")
plot(m2, which = 2, main = "M2 Q-Q")
plot(m3, which = 2, main = "M3 Q-Q")

data.frame(
  Model   = c("M1: ESU only", "M2: max_SST + ESU", "M3: Mean_SST + ESU"),
  W       = round(c(shapiro.test(residuals(m1))$statistic,
                    shapiro.test(residuals(m2))$statistic,
                    shapiro.test(residuals(m3))$statistic), 3),
  p_value = round(c(shapiro.test(residuals(m1))$p.value,
                    shapiro.test(residuals(m2))$p.value,
                    shapiro.test(residuals(m3))$p.value), 3)
) %>%
  kable(col.names = c("Model", "W", "p-value"),
        caption   = "Table 6. Shapiro-Wilk Tests of Normality",
        bootabs = TRUE) |>
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)
```


```{r mod-3e, echo = FALSE, fig.cap = "Figure 7. VIF for Model 2 and 3 (Model 1 has one predictor)."}
extract_vif <- function(model, model_name) {
  v <- vif(model)

  if (is.matrix(v)) {
    as.data.frame(v) |>
      rownames_to_column("Predictor") |>
      transmute(
        Model,
        Predictor,
        VIF = `GVIF^(1/(2*Df))`
      ) |>
      mutate(Model = model_name)
  } else {
    tibble(
      Model = model_name,
      Predictor = names(v),
      VIF = as.numeric(v)
    )
  }
}

vif_table <- bind_rows(
  extract_vif(m2, "M2"),
  extract_vif(m3, "M3")
) |>
  mutate(VIF = round(VIF, 3))

vif_table |>
  kable(
    caption  = "Table 7. VIF M2 and M3",
    booktabs = TRUE,
    align    = "llc"
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  ) |>
  row_spec(which(vif_table$VIF > 5), bold = TRUE)
```

```{r mod-3f, include=FALSE}
# Summaries
summary(m1)
summary(m2)
summary(m3)

# AIC and BIC comparison
AIC(m1, m2, m3)
BIC(m1, m2, m3)
```

Model performance was evaluated using R^2, adjusted R^2, AIC, and BIC (Table 8). The baseline model (M1: ESU only) explained little variance in mean SAR (R^2 = 0.087, Adj.R^2 = 0.056) and produced the highest AIC (57.418) and BIC (61.720) of the three models, indicating poor fit. The addition of max SST to the baseline (M2) substantially improved explanatory power (R^2 = 0.365, Adj.R^2 = 0.319), representing a four-fold increase over M1, and produced the lowest AIC (48.187) and BIC (53.923) of all models. The addition of Mean SST (M3) similarly improved fit over the baseline (R^2 = 0.310, Adj.R^2 = 0.260, AIC = 50.753, BIC = 56.489), though performance was weaker than M2 across all metrics. The AIC difference between M2 and M3 exceeded 2 units, providing meaningful evidence in favor of M2 as the most parsimonious model. Taken together, these results indicate that max SST explains meaningful variance in mean SAR beyond ESU designation alone, while Mean SST provides a weaker but still informative contribution. (Table 8)

```{r mod-3g, echo = FALSE, tbl.cap = "Table 8. Model comparisons (R^2, AIC, BIC)."}
model_comparison <- data.frame(
  Model       = c("M1: ESU only", "M2: max_SST + ESU", "M3: Mean_SST + ESU"),
  R2          = round(c(summary(m1)$r.squared, 
                        summary(m2)$r.squared, 
                        summary(m3)$r.squared), 3),
  Adj_R2      = round(c(summary(m1)$adj.r.squared, 
                        summary(m2)$adj.r.squared, 
                        summary(m3)$adj.r.squared), 3),
  AIC         = round(AIC(m1, m2, m3)$AIC, 3),
  BIC         = round(BIC(m1, m2, m3)$BIC, 3)
)

kable(model_comparison,
    col.names = c("Model", "R^2", "Adj. R^2", "AIC", "BIC"),
    caption   = "Table 8. Model comparisons (R^2, AIC, BIC)",
    booktabs = TRUE,
    align    = "llc"
  ) |>
kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE)
```

Three linear models were fit to evaluate whether sea surface temperature explains variance in mean SAR beyond ESU designation alone. The baseline model (M1), including ESU designation as the sole predictor, explained little variance in mean SAR ($R^2 = 0.087$, Table~8), and ESU failed to reach significance ($\beta_1 = 0.352$, $p = 0.107$), suggesting that designation period alone is insufficient to explain variation in smolt-to-adult return rates.

The addition of maximum sea surface temperature (M2) substantially improved model fit ($R^2 = 0.365$, adjusted $R^2 = 0.319$, Table~8), with both maximum SST ($\beta_1 = -0.273$, $p = 0.002$) and ESU ($\beta_2 = 0.640$, $p = 0.003$) reaching significance. The negative coefficient for maximum SST indicates that each one-degree Celsius increase in peak sea surface temperature is associated with a decrease of 0.273 in mean SAR, holding ESU designation constant—an ecologically meaningful result consistent with established thermal stress effects on juvenile salmon during early ocean residence.

The addition of mean SST (M3) also improved model fit relative to the baseline ($R^2 = 0.310$, adjusted $R^2 = 0.260$, Table~8), with both mean SST ($\beta_1 = -0.472$, $p = 0.006$) and ESU ($\beta_2 = 0.411$, $p = 0.038$) statistically significant, although overall model performance was weaker than that of M2 across all evaluation metrics. Scatterplots of mean SAR against each SST predictor (Figure~4) visually supported the negative linear relationships identified in models M2 and M3.

Model comparison using Akaike and Bayesian information criteria (AIC and BIC; Table~8) identified M2 as the best-fitting and most parsimonious model, with an AIC difference exceeding 2 units relative to M3 and 9 units relative to M1. Accordingly, the null hypothesis that SST explains no variance in mean SAR beyond ESU designation alone is rejected for both M2 and M3, with maximum SST identified as the stronger environmental predictor.
 




## iNterpret.

Consider including: These results are best characterized as descriptive of observed trends in the available record rather than inferential generalizations to a broader population.

Our data suggest continued support for a statistically significant flatline in the recovery of our Evolutionarily Significant Unit (ESU) despite nearly 20 years of protection under the Endangered Species Act (ESA) and critical habitat designation (4). Early models we developed show some promise (R^2= 50-90%, accounting for complexity) in explaining the linked impacts of climatic variables and food sources on the recovery of these salmonids. However, none of these models were statistically significant (p-value > 0.05). Consequently, this study was used to expand on the dataset as necessary to assess the statistical power of these findings and gain deeper insight into the ongoing challenges facing this important salmon species (1).

We expanding our analyses of data from 1994-2005 and 2006-2024, before and after the critical habitat listing for Spring-run Chinook Salmon in the Columbia River Basin (3), revealing confirmation of no significant change in the Chinook Salmon Smolt-to-Adult Ratio (SAR) further supported by no significant change in Juvenile return rates. However, there was a significant difference found in annual Maximum Sea Surface Temperature (Max SST)(2). Our prior study found a marginally significant difference in Copepod species richness along the U.S. West Coast and we hope to revisit this indicator in a future study. We also did a cursory analysis of Flooding event data plan to perform further analysis. Further data processing and modeling were conducted to explore these patterns and investigate potential relationships.

We validated our assumptions for linearity, normality, and equal variance before fitting the models. Multiple linear regressions were performed, yielding R^2 values. We also tested the models using Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC), which both indicated a good fit for the chosen models. However, the models themselves were statistically insignificant (Model 1 p-value = 0.4288; Model 2 p-value = 0.107). This suggests that increasing the sample size could improve the statistical resolution of these relationships. Although no significant patterns emerged, the alignment of R^2, AIC, and BIC results warrant further investigation (7).

By expanding our data range, we aimed to find statistically significant models compared to our prior study. Despite the limitations of annual data in studies of SAR, the importance of this variable over a 30+ year period is crucial. Although the data were drawn from a large dataset, we relied on annual means due to the limited availability of our calculated response variable, Mean SAR, for each year. Our failure to reject the null hypothesis—that there is no difference in Mean SAR between the pre- and post-listing periods—was further supported by similar findings in juvenile catches for the same species along the Oregon coastline. A follow-up study, which includes this larger range of annual data and employs a wider range of explanatory variables could potentially reveal clearer patterns and predictive models.

Obstacles Encountered in Work:

We encountered some challenges sharing dataframes in the cloud. Although the data was sourced from APIs and web scrapes, the datasets became quite large. When attempting to share with project partners, we faced permission issues that, despite the assistance of the TAs, we couldn't resolve. This highlighted an important aspect of working with Big Data, and we hope to address it using Google Cloud Storage buckets or other cloud services in the future. As a result, data processing was significantly slowed. While we managed to process data from large datasets, communication and collaboration were impacted. This led to directional changes in the project and increased independence. Ultimately, we had to abandon the Copepod and storm data in favor of a more individual approach to computation. I'm glad both team members were able to complete the assignment and use all required tools. I believe we will learn from this experience and improve our data manipulation skills.

Regarding Dataprep, after completing the data wrangling, our free subscription expired, cutting off access to all dataflows. Fortunately, the scrubbing was complete. We will need to explore how to regain access if we use Dataprep in future coursework and projects.

References:

1.  Anderson, P., Hughes, R., Team Nexus Data Wrangling Project. Oregon State University. CS512. Winter 2025. "Climate & Salmon: Assessing the Impact of Critical Habitat Designation, Climate Variables, and Food Availability on Smolt-to-Adult Return Rates for Spring-Run Chinook Salmon in Upper Columbia River Basin." Accessed 14 February 2025. https://tinyurl.com/2p8b6f3k

2. Columbia Basin Research. "DART PIT Tag Columbia Basin ESU & DPS Smolt-to-Adult Return (SAR) Survival." Columbia Basin Research, University of Washington, https://www.cbr.washington.edu/dart/query/pit_sar_esu. Accessed on 3.15.2025.

3. NOAA Coral Reef Watch. "NOAA Coral Reef Watch Operational Daily Near-Real-Time Global 5-km Satellite Coral Bleaching Monitoring Products." https://coastwatch.pfeg.noaa.gov/erddap/griddap/NOAA_DHW.html. Accessed March 13, 2025. 

4. National Centers for Environmental Information. "Storm Events Database CSV Files." NOAA, https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/. Accessed 3.12.2025.
     
5.  NOAA Fisheries. (2024, December 6). 2024 Summary of Ocean Ecosystem Indicators. Science & Data, NOAA Fisheries. Accessed 24 January 2025. <https://tinyurl.com/29wjp8za1>

6. NOAA Fisheries. (2024, August 23). Upper Columbia River Spring-run
Chinook Salmon. Endangered Species Conservation, NOAA Fisheries.
Accessed 24 January 2025. <https://tinyurl.com/y3rnr7kh>

7. NOAA Fisheries. (2024, October 28). Salmon and Steelhead Research in the
Pacific Northwest. Science & Data, NOAA Fisheries. Accessed 24 January
2025. <https://tinyurl.com/y4jhxrzp>

8. NOAA Fisheries. (2024, December 6). 2024 Summary of Ocean Ecosystem
Indicators. Science & Data, NOAA Fisheries. Accessed 24 January 2025.
<https://tinyurl.com/29wjp8za>

9. NOAA Fisheries. (2024, March 19). Oceanography of the Northern
California Current Study Area. West Coast, NOAA Fisheries. Accessed 24
January 2025. <https://tinyurl.com/yw6pj5tx>

10. Sutherland, C., Hare, D., Johnson, P. J., Linden, D. W., Montgomery, R. A., & Droge, E. (2023). Practical advice on variable selection and reporting using Akaike information criterion. Proceedings. Biological sciences, 290(2007), 20231261. https://doi.org/10.1098/rspb.2023.1261

Coding Sources:

11. scikit-learn. "KNNImputer — scikit-learn 1.6.1 documentation." https://scikit-learn.org/stable/modules/generated/sklearn.impute.KNNImputer.html. Accessed March 16, 2025.

12. Bobbitt, Z. (2022, March 31). How to Convert Pandas GroupBy Output to DataFrame. Statology. Accessed 27 January 2025. https://www.statology.org/pandas-groupby-to-dataframe/.  

13. Datetime – Basic date and time types. (n.d.). Python Standard Library. Accessed 26 January 2025. https://docs.python.org/3/library/datetime.html  

14. Ebahrim, M. (2023, December 11). Convert CSV to JSON using Python Pandas (Easy Tutorial). Like Geeks. Accessed 26 January 2025. https://likegeeks.com/csv-to-json-python-pandas/  

15. Nelamali, N. (2024, October 31). Python Pandas: Convert JSON to CSV. Spark by Examples. Accessed 26 January 2025. https://sparkbyexamples.com/pandas/python-pandas-convert-json-to-csv/  

16. Pandas API Reference. (n.d.). Pandas. Accessed 26 January 2025. https://pandas.pydata.org/docs/reference/index.html  

17. Pandas Read CSV in Python. (2024, November 21). Geeks for Geeks. Accessed 26 January 2025. https://www.geeksforgeeks.org/python-read-csv-using-pandas-read_csv/.  
18. Pykes, K. (2024, December 2). Pandas read_csv() Tutorial: Importing Data. Datacamp. Accessed 26 January 2025. https://www.datacamp.com/tutorial/pandas-read-csv  
19. pandas development team. (n.d.). pandas.DataFrame. Accessed 14 February 2025. https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.html  

20. pandas development team. (n.d.). pandas.DataFrame.drop. Accessed 14 February 2025. https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.drop.html  

21. pandas development team. (n.d.). pandas.DataFrame.transpose. Accessed 14 February 2025. https://pandas.pydata.org/pandas-docs/version/1.2/reference/api/pandas.DataFrame.transpose.html  

22. pandas development team. (n.d.). pandas.Series.str.strip. Accessed 14 February 2025. https://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.Series.str.strip.html  

23. Stack Overflow. (n.d.). What is the difference between UTF-8 and ISO-8859-1 encodings?. Accessed 14 February 2025. https://stackoverflow.com/questions/7048745/what-is-the-difference-between-utf-8-and-iso-8859-1-encodings  

24. Python Software Foundation. (n.d.). codecs — Codec registry and base classes. Accessed 14 February 2025. https://docs.python.org/3/library/codecs.html  

25. pandas development team. (n.d.). pandas.DataFrame.rename. Accessed 14 February 2025. https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.rename.html  

26. pandas development team. (n.d.). pandas.DataFrame.filter. Accessed 14 February 2025. https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.filter.html  

27. pandas development team. (n.d.). Merge, join, concatenate and compare. Accessed 14 February 2025. https://pandas.pydata.org/docs/user_guide/merging.html